---
title: "PISA 2022 Exploratory Data Analysis"
date: "2023-12-28"
editor: visual
---

## Package Installation and Data Summary:

```{r}

### Step 1:  import libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(scales)
library(RColorBrewer)
# library(ggridges)
library(knitr)
# library(kableExtra)
# library(DT)
# library(shiny)


# Step 2:  import data

# load Rdata
load("pisa.RData")
# check the data
#str(pisa)
length(pisa)
# head(pisa,10)


# drop x column
pisa <- pisa %>% select(-X)
# check summary
#summary(pisa)

#sapply(pisa, function(x) sum(is.na(x)))
# check the data

#str(pisa)
#length(pisa)


```

## Mean Of Math, Reading and Science Scores By OECD Countries:

Sums and means of PV scores for each subject and each row are summarised as additional columns:

```{r}
# Step 3:  EDA

# calculate the PV1MATH+PV2MATH+PV3MATH+PV4MATH+PV5MATH+PV6MATH+PV7MATH+PV8MATH+PV9MATH+PV10MATH and save it as a new column
# get the column names of the math scores avg
#colnames(pisa)[5:14]
pisa$SUMMATH <- rowSums(pisa[, 5:14], na.rm = TRUE)
pisa$mean_math <- rowMeans(pisa[, 5:14], na.rm = TRUE)
#calculate mean mat
# get READ scores columns
#colnames(pisa)[15:24]
# calculate the SCI score and save it as a new column
pisa$SUMREAD <- rowSums(pisa[, 15:24], na.rm = TRUE)
pisa$mean_read <- rowMeans(pisa[, 15:24], na.rm = TRUE)
# get SCI scores columns
#colnames(pisa)[25:34]
# calculate the SCI score and save it as a new column
pisa$SUMSCI <- rowSums(pisa[, 25:34], na.rm = TRUE)
pisa$mean_sci <- rowMeans(pisa[, 25:34], na.rm = TRUE)

```

Mean column is grouped by OECD countries for each subject:

```{r}
# caluculate the mean of the math, read and sci scores by OECD country
pisa_mean <- pisa %>% filter(OECD_x == 1) %>% group_by(CNT) %>% summarise(mean_math = mean(mean_math, na.rm = TRUE),
                                                  mean_read = mean(mean_read, na.rm = TRUE),
                                                  mean_sci = mean(mean_sci, na.rm = TRUE))

```

## Mean Of Math, Reading and Science Scores by Lack Of Teaching Stuff:

```{r}
# SC017Q01NA Instruction hindered by: A lack of teaching staff #Not at all # Very little # To some extent # A lot
#pisa$SC017Q01NA
# group by SC017Q01NA calculate mean math score , mean read score and mean sci score
pisa_lack_of_teaching_staff <- pisa %>% filter(OECD_x ==1) %>% group_by(SC017Q01NA) %>% summarise(mean_math = mean(mean_math, na.rm = TRUE),
                                                         mean_read = mean(mean_read, na.rm = TRUE),
                                                         mean_sci = mean(mean_sci, na.rm = TRUE))

 pisa_lack_of_teaching_staff <- pisa_lack_of_teaching_staff %>% filter(!is.na(SC017Q01NA))
# line plot
pisa_lack_of_teaching_staff %>% gather(key = "subject", value = "score", mean_math, mean_read, mean_sci) %>%
  ggplot(aes(x = SC017Q01NA, y = score, fill = subject)) +
  geom_bar(position = "dodge", stat="identity") +
  labs(title = "Mean Math, Reading and Science Score by Lack of Teaching Staff", x = "Lack of Teaching Staff",
       y = "Mean Score") + facet_grid(. ~ subject) + theme_bw() +  
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "none")

```

When there is no lack of teaching staff, the mean scores in all three subjects are interestingly lower than the cases where there is very little to some extent lack of staff. On the other hand, the mean score decreases again when there is serious lack of teaching staff as we would expect.

## Mean Math Score By Professional Development Provided To Teaching Staff:

```{r}
#  SC184*** mean math score, mean read score and mean sci score "Does your school offer professional development to
# mathematics teachers in any of the following?
#pisa$SC184Q01JA # 1Yes 2No
# set 1 as yes and 2 as no

pisa_SC184Q01JA <- pisa %>% mutate(SC184Q01JA = ifelse(SC184Q01JA == 1, "Yes", "No")) %>% filter(OECD_x == 1)
# drp the NA 
pisa_SC184Q01JA <- pisa_SC184Q01JA %>% filter(!is.na(SC184Q01JA))
professional_development_to_mathematics_teachers <- pisa_SC184Q01JA %>% group_by(SC184Q01JA) %>% summarise(mean_math = mean(mean_math, na.rm = TRUE))
# line plot normalize the data
ggplot(professional_development_to_mathematics_teachers, aes(x = SC184Q01JA, y = mean_math)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "Mean Math Score by Professional Development to Mathematics Teachers",
         x = "Professional Development to Mathematics Teachers",
         y = "Mean Math Score") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))
#pisa_SC184Q01JA$SC184Q01JA


```

The score does not seem to be changing much based on professional development or not.

Now let's have a closer look on how the trend changes based on the professional programme attendance of teaching staff in the last three month.

```{r}
#  During the last three months, what percentage of teaching
# staff in your school has attended a programme of
# professional development?
# SC025Q01NA A Staff who teach mathematics at your
# school
#pisa$SC025Q02NA
# 0-25 26-50 51-75 76-100
pisa_SC025Q02NA <- pisa %>% filter(OECD_x ==1) %>% mutate(SC025Q02NA = ifelse(SC025Q02NA <= 25, 25, ifelse(SC025Q02NA <= 50, 50, ifelse(SC025Q02NA <= 75, 75, 100))))
#pisa_SC025Q02NA$SC025Q02NA
# drop nas
pisa_SC025Q02NA <- pisa_SC025Q02NA %>% filter(!is.na(SC025Q02NA))
# group by SC025Q02NA calculate mean math score
pisa_SC025Q02NA <- pisa_SC025Q02NA %>% group_by(SC025Q02NA) %>% summarise(mean_maths = mean(mean_math, na.rm = TRUE))
# line plot normalize the data
ggplot(pisa_SC025Q02NA, aes(x = SC025Q02NA, y = mean_maths, color = "#E6E6FA")) +
    geom_line() + 
    labs(title = "Mean Math Score by Percentage of Teaching Staff Attended a Programme of Professional Development",
         x = "Percentage of Teaching Staff Attended a Programme of Professional Development",
         y = "Mean Math Score") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5)) + theme(legend.position = "none")

```

Even though the gap is not highly significant, we can talk about an upward trend as the teaching staff gets more professional development opportunities.

## Mean Math Score of OECD Countries by Public or Private School:

```{r}

# SC013 Is your school a public or a private school?
#pisa$SC013Q01TA
# group by SC013Q01TA calculate mean math score # 1Public 2Private
pisa_SC013Q01TA <- pisa %>% filter(OECD_x ==1) %>% group_by(SC013Q01TA) %>% summarise(mean_maths = mean(mean_math, na.rm = TRUE)) 
pisa_SC013Q01TA$SC013Q01TA <- ifelse(pisa_SC013Q01TA$SC013Q01TA == 1, "Public", "Private")
# drop nas
pisa_SC013Q01TA <- pisa_SC013Q01TA %>% filter(!is.na(SC013Q01TA))
#  plot  the data
ggplot(pisa_SC013Q01TA, aes(x = SC013Q01TA, y = mean_maths)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = "Mean Math Score by Public or Private School",
         x = "Public or Private School",
         y = "Mean Math Score") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5))


```

```{r}

# SC013 Is your school a public or a private school?
#pisa$SC013Q01TA
# group by SC013Q01TA calculate mean math score # 1Public 2Private
pisa_SC013Q01TA <- pisa %>% group_by(SC013Q01TA) %>% summarise(mean_math, na.rm = TRUE) 
pisa_SC013Q01TA$SC013Q01TA <- ifelse(pisa_SC013Q01TA$SC013Q01TA == 1, "Public", "Private")
# drop nas
pisa_SC013Q01TA <- pisa_SC013Q01TA %>% filter(!is.na(SC013Q01TA))
#group by countries
pisa_SC013Q01TA <- pisa_SC013Q01TA %>% filter(!is.na(SC013Q01TA))
pisa_SC013Q01TA <- pisa_SC013Q01TA 
#  plot  the data

ggplot(pisa_SC013Q01TA, aes(x = mean_math)) +
geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 100) +
  stat_bin(
    binwidth = 100,
    aes(y = after_stat(count / sum(count)), 
        label = scales::percent(after_stat(count / sum(count))))) +
  scale_y_continuous(labels = scales::percent) +    facet_grid(. ~ SC013Q01TA) + labs(title = "Mean Math Score by Public or Private School",
         x = "Mean Math Score") 


```

```{r}

pisa_SC013Q01TA <- pisa %>% group_by(SC013Q01TA) %>% summarise(mean_math, na.rm = TRUE) 
pisa_SC013Q01TA$SC013Q01TA <- ifelse(pisa_SC013Q01TA$SC013Q01TA == 1, "Public", "Private")
# drop nas
pisa_SC013Q01TA <- pisa_SC013Q01TA %>% filter(!is.na(SC013Q01TA))

#  plot  the data
ggplot(data = pisa_SC013Q01TA, aes(mean_math)) + 
  geom_histogram(bins = 100,
                 aes(y = stat(100*density))) +
  facet_wrap(.~SC013Q01TA) +
  scale_y_continuous(labels = percent_format()) +
  theme_bw()
```

```{r}

pisa_SC013Q01TA <- pisa %>% filter(CNT =="TUR") %>% group_by(SC013Q01TA) %>% summarise(mean_math, na.rm = TRUE) 
pisa_SC013Q01TA$SC013Q01TA <- ifelse(pisa_SC013Q01TA$SC013Q01TA == 1, "Public", "Private")
# drop nas
pisa_SC013Q01TA <- pisa_SC013Q01TA %>% filter(!is.na(SC013Q01TA))
 
#  plot  the data
ggplot(data = pisa_SC013Q01TA, aes(mean_math)) + 
  geom_histogram(bins = 10,
                 aes(y = stat(10*density))) +
  facet_wrap(.~SC013Q01TA) +
  scale_y_continuous(labels = percent_format()) +
  theme_bw()
```

```{r}

# SC013 Is your school a public or a private school?
#pisa$SC013Q01TA
# group by SC013Q01TA filter TR calculate mean math score # 1Public 2Private
pisa_SC013Q01TA <- pisa %>% filter(CNT=="TUR") %>% group_by(SC013Q01TA) %>% summarise(mean_math, na.rm = TRUE) 
pisa_SC013Q01TA$SC013Q01TA <- ifelse(pisa_SC013Q01TA$SC013Q01TA == 1, "Public", "Private")
# filter public school
pisa_SC013Q01TA_public <- pisa_SC013Q01TA %>% filter(SC013Q01TA == "Public")

#  plot  the data

ggplot(pisa_SC013Q01TA_public, aes(x = mean_math)) +
geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 100) +
  stat_bin(
    binwidth = 100, fill="#ECD5C4",
    aes(y = after_stat(count / sum(count)), 
        label = scales::percent(after_stat(count / sum(count))))) +
  scale_y_continuous(labels = scales::percent)  + labs(title = "Mean Math Score by Public School",
         x = "Mean Math Score") 


```

```{r}


# SC013 Is your school a public or a private school?
#pisa$SC013Q01TA
# group by SC013Q01TA calculate mean math score # 1Public 2Private
pisa_SC013Q01TA <- pisa %>% filter(CNT=="TUR") %>% group_by(SC013Q01TA) %>% summarise(mean_math, na.rm = TRUE) 
pisa_SC013Q01TA$SC013Q01TA <- ifelse(pisa_SC013Q01TA$SC013Q01TA == 1, "Public", "Private")
# drop nas
pisa_SC013Q01TA_private <- pisa_SC013Q01TA %>% filter(SC013Q01TA == "Private")

#  plot  the data

ggplot(pisa_SC013Q01TA_private, aes(x = mean_math)) +
geom_histogram(aes(y = after_stat(count / sum(count))), binwidth = 100) +
  stat_bin(
    binwidth = 100, fill = "#D8BFD8",
    aes(y = after_stat(count / sum(count)), 
        label = scales::percent(after_stat(count / sum(count))))) +
  scale_y_continuous(labels = scales::percent)  + labs(title = "Mean Math Score by Private School",
         x = "Mean Math Score", y = "Percentage") 


```

\`\`\`
